Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-02-20 appendix.tex
add step-b-step linearizability, not complete this section

\DataDifferentiatedisEnoughforPQ

@@

@@

The following lemma states that $\textit{PQueue}$ is closed under projection. This lemma holds since the predicates appears in inductive rules of $\textit{PQueue}$ are closed under projection.

\begin{restatable}{lemma}{PQisClosedUnderProjection}
\label{lemma:PQ is closed under projection}

$\textit{PQueue}$ is closed under projection.

%When the context is clear, we shorten priority-value data-independence as data-independence.
For priority queue, the notion of step-by-step linearizability is modified as follows:

%\vspace{-6pt}
\begin{definition}[step-by-step linearizability for priority queue]\label{def:step-by-step linearizability of priority queue}
A data structure $S=R_1, \ldots, R_n$ is said be to step-by-step linearizable, if for any data-differentiated sequence $e$, if $e$ is linearizable w.r.t. $\textit{MR}_{\textit{pri}}$ ($1 \leq i \leq 4$) with witness $x$, we have:

$e \setminus x \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket \Rightarrow e \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$

\noindent if $e$ is linearizable w.r.t. $\textit{MR}_{\textit{pr5}}$ with witness $x$, we have:

$e \setminus x \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr5}} \rrbracket \Rightarrow e \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr5}} \rrbracket$
\end{definition}

Let $\textit{proj}(s)$ denote the projection of $s$ obtained by erasing all the operations of several items from $s$. Here we assume that each $\textit{rm}(\textit{empty})$ is differentiated from each other, and we can consider each $\textit{rm}(\textit{empty})$ as having a unique item.

\begin{restatable}{lemma}{PQasMultiInMRpriforSequence}
\label{lemma:PQ as multi in MRpri for sequence}
$s \in \textit{PQueue}$, if and only if, $\forall s'$, if $s' \in \textit{proj}(s)$, then $s' \in \textit{MR}_{\textit{pri}}$ for some $0 \leq i \leq 5$.
\end{restatable}

To prove the $\textit{if}$ direction. The $\textit{if}$ direction holds trivially when $s = \epsilon$. When $s \neq \epsilon$. since $s \in \textit{proj}(s)$, we generate sequence $s_1$ from $s$ as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $s$ has $\textit{rm}(\textit{empty})$ operations: $s_1$ is generated from $s$ by erasing some $\textit{rm}(\textit{empty})$ operation.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in $S_1$ and unmatched $\textit{put}$ operations of some items in $S_2$: $s_1$ is generated from $s$ by erasing the unmatched $\textit{put}$ operation of items last putted in $S_2$.

\item[-] If the maximal priority in $s$ is unmatched $\textit{put}$ operations of some items in $S_2$ and $\vert S_2 \vert =1$: $s_1$ is generated from $s$ by erasing this unmatched $\textit{put}$ operation.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert \geq 2$: $s_1$ is generated from $s$ by erasing matched $\textit{put}$ and $\textit{rm}$ operations of the item which is last putted in $S$.

\item[-] If the maximal priority in $s$ is of matched $\textit{put}$ and $\textit{rm}$ operations of some items in set $S$ and $\vert S \vert = 1$: $s_1$ is generated from $s$ by erasing matched $\textit{put}$ and $\textit{rm}$ operations of the item in $S$.
\end{itemize}

Similarly, we obtain $s_2$ from $s_1$, obtain $s_3$ from $s_2$, $\ldots$, until we obtain some $s_m = \epsilon$. It is obvious that $s_m \in \textit{PQueue}$, and by the five rules of $\textit{PQueue}$, it is easy to see that $s_{\textit{m-1}},\ldots,s_1,s \in \textit{PQueue}$. \qed
\end {proof}

@@

\begin{restatable}{lemma}{Rpr1isStepByStepLinearizability}
\label{lemma:Rpr1 is step-by-step linearizability}

$R_{\textit{pr1}}$ is step-by-step linearizability.

Assume that $l$ matches $R_{\textit{pr1}}$ with witness $x$. Assume that  Let $h$ be a data-differentiated history, and $l$ be an operation sequence such that $h \sqsubseteq l$. Then, let $h'=h \setminus x$ and assume that $h' \sqsubseteq \llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$. Let $l'$ be a sequence in $\llbracket R_{\textit{pr0}},\ldots,R_{\textit{pr4}} \rrbracket$ such that $h' \sqsubseteq l'$.

According to $R_{\textit{pr1}}$, there exist sequences $u$, $v$, and $w$, such that $l=u \cdot \textit{put}(x,\textit{pri}_x) \cdot v \cdot \textit{rm}(x) \cdot w$, where the $\textit{put}$ operations and $\textit{rm}$ operations in $u \cdot v$ are matched, and $\textit{pri}_x$ is larger than any priority in $u \cdot v \cdot w$. Let $D_L$ be the set of items in $u \cdot v$, let $D_R$ be the set of items in $w$. Since $u \cdot v$ contain only matched $\textit{put}$ and $\textit{rm}$ operations, we can see that $D_L \cap D_R = \emptyset$. Let $l'_L = l' \vert_{D_L}$ and $l'_R = l' \vert_{D_R}$. Let $S_1$ be the set of all operations of $h$ that happen before $\textit{put}(x,\textit{pri}_x)$, let $l'_{L1}$ be the shortest prefix of $l'_L$ that contains all operations of $S_1$, and let $l'_{L2}$ be the sequence such that $l'_L = l'_{L1} \cdot l'_{L2}$. It is obvious that if $\textit{put}(x,\textit{pri}_x)$ happens before an operation $o$ in $h$, then $o$ is in $l'_{L2}$. Let sequence $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R$.

Since priority queue is closed under projection (Lemma \ref{lemma:PQ is closed under projection}) and $u \cdot v$ contain matched $\textit{put}$ and $\textit{rm}$ operations, we know that $l'_{L1} \cdot l'_{L2} \in \textit{PQueue}$ and $l'_R \in \textit{PQueue}$. Since $\textit{pri}_x$ is larger than any priority in $h$, it is easy to see that $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$. Since the priority queue is empty after executing $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$, we know that $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R \in \textit{PQueue}$.

@@

\begin{restatable}{lemma}{PQueueisStepByStepLinearizability}
\label{lemma:PQueue is step-by-step linearizability}

$\textit{PQueue}$ is step-by-step linearizability.

>>>>

\DataDifferentiatedisEnoughforPQ*

@@

\section{Proofs in Section \ref{sec:step-by-step linearizability of priority queues}}
\label{sec:appendix in section step-by-step linearizability of priority queues}

\PQisClosedUnderProjection*

This is obvious, since for each conditions in the $\textit{Guard}$ part of the rules of priority queue, if a sequence of sequential executions satisfy it, then its sub-sequence also satisfy it. For example, if $\textit{rm}(\textit{empty}) \notin l = u \cdot v \cdot w$ and let $D_l$ be the set of items of $l$, then for each subset $D' \subseteq D$, it is obvious that $\textit{rm}(\textit{empty}) \notin l \vert_{ D' }$. Similar cases hold for $\textit{pri} \geq \textit{maxPri}(l)$, $\textit{pri} > \textit{maxUPri}(l)$, $\textit{matched}(l)$, $\textit{putInSeq}(l,l_1,\textit{pri})$ (let $l_1$ be sub-sequence of $l$). This completes the proof of this lemma. \qed
\end {proof}

\PQasMultiInMRpriforSequence*

The \textit{only if} direction is a direct consequence of Lemma \ref{lemma:PQ is closed under projection}, since it is easy to see that each sequence in $\textit{PQueue}$ also belongs to $\textit{MS}(R)$.

To prove the $\textit{if}$ direction. The $\textit{if}$ direction holds trivially if $e = \epsilon$. When $e \neq \epsilon$, we generate sequence $e_1$ from $e$ as follows:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] If $\textit{last}(e) = \textit{PQ}_3$: $e_1$ is generated from $e$ by erasing one $\textit{rm}(\textit{empty})$.

\item[-] Else, if $\textit{last}(e) = \textit{PQ}_2^{=}$, and the maximal priority of $e$ is unmatched $\textit{put}$ with items in $S$ and matched $\textit{put}$: $e_1$ is generated from $e$ by erasing one unmatched $\textit{put}$ which use the item last putted in $S$.

\item[-] Else, if $\textit{last}(e) = \textit{PQ}_2^{>}$, and the maximal priority in $s$ is unmatched $\textit{put}$ with items in $S$: $e_1$ is generated from $e$ by erasing one unmatched $\textit{put}$ which use the item last putted in $S$.

\item[-] Else, if $\textit{last}(e) = \textit{PQ}_1^{=}$, and the maximal priority in $e$ is of more than one pair of matched $\textit{put}$ with items in $S$: $e_1$ is generated from $e$ by erasing matched $\textit{put}$ and $\textit{rm}$ of the item which is last putted in $S$.

\item[-] Else, if $\textit{last}(e) = \textit{PQ}_1^{>}$: $e_1$ is generated from $e$ by erasing a pair of matched $\textit{put}$ and $\textit{rm}$ with maximal priority in $e$.
\end{itemize}

Similarly, we obtain $e_2$ from $e_1$, obtain $e_3$ from $e_2$, $\ldots$, until we obtain some $e_m = \epsilon$. Let $e_0=e$. It is obvious that $e_m \in \textit{PQueue}$. For $e_{\textit{m-1}}$, since

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $e_m \in \textit{PQueue}$,

\item[-] By assumption, we know that $e_{\textit{m-1}} \in \textit{MS}(R_{\textit{m-1}})$, where $R_{\textit{m-1}} = \textit{last}(R_{\textit{m-1}})$. This implies that the guard of $R_{\textit{m-1}}$ is satisfied.
\end{itemize}

Therefore, we know that $e_{\textit{m-1}} \in \textit{PQueue}$. Similarly, we can prove that $e_{\textit{m-2}},\ldots,e_0 = e \in \textit{PQueue}$. \qed
\end {proof}

The prove that $\textit{PQueue}$ is step-by-step linearizability, we investigate each rules individually.

Given a data-differentiated execution and its history, we can abuse notation and mix labels and method events with operations themselves, since items are unique in a data-differentiated execution. For instance, we will reference an operation labeled by $\textit{put}(p,a)$ as $\textit{put}(p,a)$.

Given an operation $o$ with call action $\textit{cal}_o (\textit{put},a,p)$ and return actions $\textit{ret}_o (\textit{put})$, its method event is $\textit{put}(a,p)$. Given an operation $o$ with call action $\textit{cal}_o (\textit{rm})$ and return actions $\textit{ret}_o (\textit{rm},a)$, its method event is $\textit{rm}(a)$.

\begin{restatable}{lemma}{PQ1LarisStepByStepLinearizability}
\label{lemma:PQ1Lar is step-by-step linearizability}
If concurrent execution $e$ is linearizable w.r.t. $\textit{MS}(\textit{PQ}_1^{>})$ with witness $x$, then $e \setminus x \sqsubseteq \textit{PQueue} \Rightarrow e \sqsubseteq \textit{PQueue}$.
\end{restatable}

Let $h$ be the data-differentiated history of $e$, and $l$ be an sequential execution such that $h \sqsubseteq l$ and $l$ matches $\textit{PQ}_1^{>}$ with witness $x$. Let $h'=h \setminus x$ and assume that $h' \sqsubseteq l' \in \textit{PQueue}$.

According to $\textit{MS}(\textit{PQ}_1^{>})$, there exist sequences $u$, $v$, and $w$, such that $l=u \cdot \textit{put}(x,\textit{pri}_x) \cdot v \cdot \textit{rm}(x) \cdot w$, where the $\textit{put}$ and $\textit{rm}$ in $u \cdot v$ are matched, and $\textit{pri}_x$ is larger than priority of any method event in $u \cdot v \cdot w$. Let $D_L$ be the set of items in $u \cdot v$, let $D_R$ be the set of items in $w$. Since $u \cdot v$ contain only matched $\textit{put}$ and $\textit{rm}$, we can see that $D_L \cap D_R = \emptyset$. Let $l'_L = l' \vert_{D_L}$ and $l'_R = l' \vert_{D_R}$. Let $S_1$ be the set of method events of operations of $h$ that happen before $\textit{put}(x,\textit{pri}_x)$, let $l'_{L1}$ be the shortest prefix of $l'_L$ that contains all method events in $S_1$, and let $l'_{L2}$ be the sequence such that $l'_L = l'_{L1} \cdot l'_{L2}$. It is obvious that if $\textit{put}(x,\textit{pri}_x)$ happens before an operation $o$ in $h$, then $o$ is in $l'_{L2}$. Let sequence $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R$.

Since priority queue is closed under projection (Lemma \ref{lemma:PQ is closed under projection}) and $u \cdot v$ contain matched $\textit{put}$ and $\textit{rm}$, we know that $l'_{L1} \cdot l'_{L2} \in \textit{PQueue}$ and $l'_R \in \textit{PQueue}$. Since $\textit{pri}_x$ is larger than any priority in $h$, it is easy to see that $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$. Since the priority queue is empty after executing $l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \in \textit{PQueue}$, we know that $l'' = l'_{L1} \cdot \textit{put}(x,\textit{pri}_x) \cdot l'_{L2} \cdot \textit{rm}(x) \cdot L'_R \in \textit{PQueue}$.

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] if $e$ is linearizable w.r.t. $\textit{MS}_{\textit{PQ}_1^{>}}$ ($R \in \{ \textit{PQ}_1^{>}, \textit{PQ}_1^{=}, \textit{PQ}_2^{>},\textit{PQ}_2^{=} \}$) with witness $x$, we have: $e \setminus x \sqsubseteq \textit{PQueue} \Rightarrow e \sqsubseteq \textit{PQueue}$.

\item[-] if $e$ is linearizable w.r.t. $\textit{MS}(\textit{PQ}_3)$ and $o$ is a $\textit{rm}(\textit{empty})$ event, we have:

$e \setminus o \sqsubseteq \textit{PQueue} \Rightarrow e \sqsubseteq \textit{PQueue}$.
\end{itemize}

\PQueueisStepByStepLinearizability*

@@

@@

@@
